Bird beak's elaboration and exploration: data preparation

### 2021-08-27 Analysis/01-Data_preparation.Rmd
analyses up
Update 01-Data_preparation.Rmd

@@

```{r load_beer, eval = TRUE}
## Loading beer
library(beer)

## Loading the demo data
data(demo_data)
data(demo_trees)
```

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Demo data

This is what the demo data looks like

```{r, plot_the_data}
## Plotting it
colour_vector <- c("orange", "blue", "darkgreen")
plot.space(demo_data,
           col = colour_vector,
           levels = demo_data$clade,
           xlab = "PC1 (90.5%)",
           ylab = "PC1 (6.86%)")
legend("topleft", legend = levels(demo_data$clade), col = colour_vector, pch = 19)
```

>>>>

The method described in the next three markdown is essentially a mini chains with pseudo-metropolis coupling markov chain generalised linear mixed model method to determine clades and tips elaboration and exploration (`mcmcpmcmcmcglmmmctee`) but we will refrain using this acronym and method name.

@@

```{r load_beer, eval = TRUE}
## Loading packages
library(dispRity)
library(beer)
library(knitr)
```

@@

 * The trait space

```{r}
## Reading the different trait spaces
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")
shapespace <- shapespace$PCscores
formspace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_FORMSPACE_FULL.rds")
formspace <- formspace$PCscores
```

```{r, echo = FALSE, fig.height = 16, fig.width = 8}
par(mfrow = c(2,1))
plot.dispRity(shapespace, type = "preview", pch = 21, main = "Shape space")
plot.dispRity(formspace, type = "preview", pch = 21, main = "Form space")
```

 * The tree(s)

```{r}
## Reading the consensus tree
consensus_tree <- read.nexus("../Data/Raw/Tree_data/9993taxa_1_tree_Global_MCC_CA.tre")

## Selecting only the tips in the trait space
matching_tips <- consensus_tree$tip.label %in% rownames(shapespace)
consensus_tree <- drop.tip(consensus_tree, tip = consensus_tree$tip.label[!matching_tips])
```

 * The clade data

```{r}
## Reading the clades list
clades_list <- as.data.frame(read.csv("../Data/Raw/BirdClades.csv", row.names = 1, stringsAsFactors = TRUE))
## Remove the absent species
clades_list <- clades_list[rownames(clades_list) %in% rownames(shapespace), ]

#@param table the clade list data frame
#@param level the ID of the column to use as a level
#@param threshold drop levels with less than the threshold
make.level.list <- function(table, level, threshold = 20) {
  ## Select the level
  table_level <- table[, level]
  ## Remove the empties
  table_level <- as.factor(as.character(table_level[-which(table_level == "")]))
  ## Select the levels below the threshold
  remove_levels <- which(table(table_level) < threshold)
  ## Remove the levels
  while(length(remove_levels) > 0) {
    table_level <- as.factor(as.character(table_level[-which(table_level == names(remove_levels)[1])]))
    remove_levels <- remove_levels[-1]
  }
  ## Convert into a list
  factors <- levels(table_level)
  levels_list <- list()
  while(length(factors) > 0) {
    levels_list[[length(levels_list) + 1]] <- which(table_level == factors[1])
    factors <- factors[-1]
  }
  ## Name the elements correctly
  names(levels_list) <- levels(table_level)
  return(levels_list)
}

## Reading the clades list
groups_level1 <- make.level.list(clades_list, level = 1, threshold = 20)
groups_level2 <- make.level.list(clades_list, level = 2, threshold = 20)
all_groups <- c(groups_level1, groups_level2)
```

### Selecting the number of axes

How many dimensions do we need to represent at least 95% of the variance in each group

```{r}
## Make a dispRity object
shapespace <- custom.subsets(shapespace, group = all_groups)
formspace  <- custom.subsets(formspace, group = all_groups)
shape_axes <- select.axes(shapespace)
form_axes  <- select.axes(formspace)
```

```{r, echo = FALSE}
kable(summary(shape_axes)[, c(1:((length(shape_axes$dimensions)+1)*2))], caption = "variance per axes for the Shape space", label = "Table X1")
```

```{r, echo = FALSE}
kable(summary(form_axes)[, c(1:((length(form_axes$dimensions)+1)*2))], caption = "variance per axes for the Form space", label = "Table X2")
```

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
plot(shape_axes)
```
Fig X1: variance per axes for the Shape space

```{r, echo = FALSE, fig.height = 18, fig.width = 18}
plot(form_axes)
```
Fig X1: variance per axes for the Form space

By analysing the distribution of the variance within each group, we will need `r length(shape_axes$dimensions)` dimensions for the shape space and `r length(form_axes$dimensions)` for the form space.

### Saving the data

```{r}
## Save the data to be analysed in Data/Processed
form_shape_space_1tree_2levels <- list(
    "consensus_tree" = consensus_tree,
    "shape_space"    = shapespace,
    "form_space"     = formspace,
    "groups_list"    = all_groups)
save(form_shape_space_1tree_2levels, file = "../Data/Processed/form_shape_space_1tree_2levels.rda")
```
